Unveiling Vulnerabilities

### 2024-04-24 paper/paper.qmd
Updated paper and qmd

---
title: "Unveiling Vulnerabilities"
subtitle: "An Analysis of Causes of Death Across Age Groups and Genders Within Toronto's Homelessness Population"
author: "Faiza Imam"
thanks: "Code and data are available at: https://github.com/fazim07/deaths_gender"
date: today
date-format: long
abstract: "This paper aims to explore the relationship between male homelessness and mortality rates across age groups, focusing on causes of death related to homelessness in Toronto from the years 2017- 2023. The hypothesis suggests that younger homeless males (<20 and 20-39) may have higher mortality rates due to harsh living conditions, limited healthcare access, and increased vulnerability. The results indicate that while male homeless mortality rates are generally higher than reported female death tolls, the most affected age demographic is the 40-59 age range rather than the hypothesized age group."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Some of the key factors that result in homelessness are financial challenges, results from abuse or intimate partner violence, health issues which cannot be sustained within a household, and finally addiction or substance abuse (@statCan). Although many underlying factors contribute to an individual becoming or facing homelessness, it’s important to keep an open and empathetic  mind when dealing with the data collected to understand these trends. Research has shown that women make up 27.3% of the homeless population. Several factors contribute to the higher likelihood of women experiencing homelessness compared to men. These factors include unstable employment, which raises the risk of income fluctuations or job loss, as well as the added responsibility of care giving, such as caring for children and dependents, intimate partner violence and other forms of abuse (@Homehubs). However, there is no study explaining the reason to the high volume of male victims in homelessness populations.

@@

When examining the figures provided in @fig-agedeathbyyear, a clear picture emerges regarding the prevalent causes of death among various age demographics over the years. One notable trend is the consistent reporting of the highest death tolls among the 40-59 age demographic across different years. In particular, the data reveals that in 2019 @agedeathbyyear (c ), this age group experienced the highest number of deaths in seven out of the nine cause-of-death categories considered. This finding underscores the significant impact of mortality within the 40-59 age range and suggests a need for targeted interventions and health strategies to address the specific challenges faced by individuals in this demographic. Further analysis of these trends can offer valuable insights into the dynamics of mortality rates among different age groups, guiding efforts to improve healthcare access, address underlying health issues, and enhance support systems for vulnerable populations, including those experiencing homelessness.

@@

## Weaknesses & Next Steps
When considering weaknesses and potential biases in research on the correlation between gender and age within the homeless population in Toronto, several aspects should be addressed to understand what factors may have skewed the data projections during this paper's analysis. There was limited data available regarding age and gender breakdowns, and even with the available data, there were a few entries that were unknown for the age category. I opted to exclude those values to focus on the existing and available values within the dataset. This decision could have skewed the data and constrained the accuracy of the findings. As mentioned previously, there is an underlying stigma surrounding homelessness, which alone can influence individuals who are reluctant to disclose information, particularly those facing hidden homelessness. Lastly, within the gender values, there are three variables: male, female, and unknown. This unknown entry possibly refers to those individuals who have an LGBTQ+ identity and can overlook important nuances in the experiences of homeless individuals. Addressing these weaknesses and biases through rigorous study design, comprehensive data collection methods, ethical considerations, and thoughtful analysis can enhance the reliability and validity of research findings in this area."

>>>>

---
title: "Unveiling Vulnerabilities"
subtitle: "An Analysis of Causes of Death Across Age Groups and Genders Within Toronto's Homelessness Population"
author: "Faiza Imam"
thanks: "Code and data are available at: https://github.com/fazim07/deaths_gender"
date: today
date-format: long
abstract: "This paper aims to explore the relationship between male homelessness and mortality rates across age groups, focusing on causes of death related to homelessness in Toronto from the years 2017- 2023. The hypothesis suggests that younger homeless males (<20 and 20-39) may have higher mortality rates due to harsh living conditions, limited healthcare access, and increased vulnerability. The results indicate that while male homeless mortality rates are generally higher than reported female death tolls, the most affected age demographic is the 40-59 age range rather than the hypothesized age group."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

Some of the key factors that result in homelessness are financial challenges, results from abuse or intimate partner violence, health issues which cannot be sustained within a household, and finally addiction or substance abuse [@statCan]. Although many underlying factors contribute to an individual becoming or facing homelessness, it’s important to keep an open and empathetic  mind when dealing with the data collected to understand these trends. Research has shown that women make up 27.3% of the homeless population. Several factors contribute to the higher likelihood of women experiencing homelessness compared to men. These factors include unstable employment, which raises the risk of income fluctuations or job loss, as well as the added responsibility of care giving, such as caring for children and dependents, intimate partner violence and other forms of abuse [@Homehubs]. However, there is no study explaining the reason to the high volume of male victims in homelessness populations.

@@

When examining the figures provided in @fig-agedeathsbyyear, a clear picture emerges regarding the prevalent causes of death among various age demographics over the years. One notable trend is the consistent reporting of the highest death tolls among the 40-59 age demographic across different years. In particular, the data reveals that in 2019 @fig-agedeathsbyyear (c), this age group experienced the highest number of deaths in seven out of the nine cause-of-death categories considered. This finding underscores the significant impact of mortality within the 40-59 age range and suggests a need for targeted interventions and health strategies to address the specific challenges faced by individuals in this demographic. Further analysis of these trends can offer valuable insights into the dynamics of mortality rates among different age groups, guiding efforts to improve healthcare access, address underlying health issues, and enhance support systems for vulnerable populations, including those experiencing homelessness.

@@

# Weaknesses & Next Steps
When considering weaknesses and potential biases in research on the correlation between gender and age within the homeless population in Toronto, several aspects should be addressed to understand what factors may have skewed the data projections during this paper's analysis. There was limited data available regarding age and gender breakdowns, and even with the available data, there were a few entries that were unknown for the age category. I opted to exclude those values to focus on the existing and available values within the dataset. This decision could have skewed the data and constrained the accuracy of the findings. As mentioned previously, there is an underlying stigma surrounding homelessness, which alone can influence individuals who are reluctant to disclose information, particularly those facing hidden homelessness. Lastly, within the gender values, there are three variables: male, female, and unknown. This unknown entry possibly refers to those individuals who have an LGBTQ+ identity and can overlook important nuances in the experiences of homeless individuals. Addressing these weaknesses and biases through rigorous study design, comprehensive data collection methods, ethical considerations, and thoughtful analysis can enhance the reliability and validity of research findings in this area."
